---
title: "Taxonomy Investigations"
format: html
---

### Library Load & Setup

```{r}
#| label: Setup
# ---
#source(here::here("EDA", "Kaitlyn_EDA", "Wet_Runs.R"))
source(here::here("EDA", "project_settings.R"))
#source(here::here("EDA", "Kaitlyn_EDA", "Reusable_Scripts", "Utility_Functions.R"))
#variable <- read.csv("~/data/courses/class_lsc585spring2026_ndixit5/Therayess/")

suppressPackageStartupMessages({
  library(ggplot2)
  library(knitr)
  library(patchwork)
  library(mia)  # for mergeData
  library(SummarizedExperiment)
})
```

# Structural Metadata Overview

## Tables

```{r}
#| label: Metadata Overview - Tables
# ---

# Single Study, nested list (Example)
Thomas2018a_taxa <- CRC_progression_studies[["ThomasAM_2018a"]][["relative_abundance"]]

# All studies combined to one list
taxa <- merge_studies(CRC_progression_studies, "relative_abundance")

# Metadata into dataframes from nested list
Thomas2018a_taxa_meta <- as.data.frame(colData(Thomas2018a_taxa))
taxa_meta <- as.data.frame(colData(taxa))

# Overview Tables
Thomas2018a_taxa_meta_overview <- eda_overview_table(Thomas2018a_taxa_meta)
taxa_meta_overview <- eda_overview_table(taxa_meta)

Thomas2018a_taxa_meta_overview %>%
  arrange(Percent_Missing) %>%
  mutate(Index = row_number()) %>%              
  select(Index, everything()) %>%               
  kbl(
    col.names = c("Index",
                  "Variable Name",
                  "Total Rows",
                  "Empty Rows",
                  "Percent Missing",
                  "Unique Values",
                  "Variable Type"),
    caption = "Structural Metadata Summary for ThomasAM_2018a Relative Abundance Data",
    format = "html"
  ) %>%
  kable_styling(full_width = FALSE) %>%
  column_spec(2, bold = TRUE)

taxa_meta_overview %>%
  arrange(Percent_Missing) %>%
  mutate(Index = row_number()) %>%              
  select(Index, everything()) %>%               
  kbl(
    col.names = c("Index",
                  "Variable Name",
                  "Total Rows",
                  "Empty Rows",
                  "Percent Missing",
                  "Unique Values",
                  "Variable Type"),
    caption = "Structural Metadata Summary for All Relative Abundance Data",
    format = "html"
  ) %>%
  kable_styling(full_width = FALSE) %>%
  column_spec(2, bold = TRUE)
```

## Plots/Maps

```{r}
#| label: Metadata Overview - Plots/Maps
# ---

# Heatmap of Metadata Completeness
taxa_meta_overview %>%
  arrange(desc(Percent_Missing)) %>%
  mutate(Column_Name = factor(Column_Name, levels = Column_Name)) %>%  # lock order
  ggplot(aes(x = "", y = Column_Name, fill = Percent_Missing)) +
  geom_tile(color = "white") +
  scale_fill_gradient(
    low = "steelblue",
    high = "white",
    limits = c(0, 100)
  ) +
  scale_y_discrete(expand = expansion(mult = c(0.02, 0.02))) +
  labs(
    title = "Structural Metadata Completeness:\nAll Studies - Relative Abundance Data",
    subtitle = "Percent rows with missing values",
    x = NULL,
    y = NULL
  ) +
  theme_bw() +
  theme(
    axis.text.y = element_text(size = 7, lineheight = 2),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    plot.title = element_text(face = "bold", size = 10, hjust = 0.5),
    plot.subtitle = element_text(size = 9, hjust = 0.5)
  )

Thomas2018a_taxa_meta_overview %>%
  arrange(desc(Percent_Missing)) %>%
  mutate(Column_Name = factor(Column_Name, levels = Column_Name)) %>%  # lock order
  ggplot(aes(x = "", y = Column_Name, fill = Percent_Missing)) +
  geom_tile(color = "white") +
  scale_fill_gradient(
    low = "steelblue",
    high = "white",
    limits = c(0, 100)
  ) +
  scale_y_discrete(expand = expansion(mult = c(0.02, 0.02))) +
  labs(
    title = "Structural Metadata Completeness:\nThomasAM_2018a - Relative Abundance Data",
    subtitle = "Percent rows with missing values",
    x = NULL,
    y = NULL
  ) +
  theme_bw() +
  theme(
    axis.text.y = element_text(size = 7, lineheight = 2),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    plot.title = element_text(face = "bold", size = 10, hjust = 0.5),
    plot.subtitle = element_text(size = 9, hjust = 0.5)
  )
```

# CRC Classifiers Spread

The spread of the CRC Progression classifiers, as well as the spread of other categorical variables across each of these categories can indicate potential sources of bias in either the data, the disease, or other sources.

## Spread of the classifiers

Given the spread, we will need to find a way to deal with the imbalance in classifier levels, particularly for the "-m" categories.

-   potential cutoff of 30? IE if there's not at least 30 to a group, add it to the next logical group (-M would go to CRC/PA pure)

-   Weighed importance for those groups with less individuals?

```{r}
#| label: CRC Progression Classifier Bias Checks
# ---
taxa_meta %>% 
  ggplot(aes(x=disease_class, stat = "count", fill = disease_class)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = after_stat(count)), 
            vjust = -0.21, size = 3.5, color = "black") +
  scale_fill_manual(values = progression_colors, 
                    name = progression_legend_title) +
  # Labeling
  #---------
  labs(y = "Sample Count", x = "CRC Progression Classes") +
  ggtitle("Counts of CRC Progression Classes in Study")
```

## Potential Biases in CRC Progression Classes

Biases can be found through proportions checking– are a higher proportion of one group in a particular classifier than another?

### Demographic and Clinical Sources of Bias

```{r}
#| label: Demographic and Clinical Characteristics
# ---
# Gender
taxa_meta %>% 
  ggplot(aes(x=disease_class, stat = "count", fill = gender)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = fareblind15) +
  # Labeling
  #---------
  labs(y = "Proportion", x = "CRC Progression Classes") +
  ggtitle("Proportion of Gender per CRC Progression Class")

# Age Category
taxa_meta %>% 
  ggplot(aes(x=disease_class, stat = "count", fill = age_category)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = fareblind15) +
  # Labeling
  #---------
  labs(y = "Proportion", x = "CRC Progression Classes") +
  ggtitle("Proportion of Age Category per CRC Progression Class")

# Age Decade
taxa_meta %>% 
  ggplot(aes(x=disease_class, stat = "count", fill = age_decade)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = fareblind15) +
  # Labeling
  #---------
  labs(y = "Proportion", x = "CRC Progression Classes") +
  ggtitle("Proportion of Age Decade per CRC Progression Class")

# Country
taxa_meta %>% 
  ggplot(aes(x=disease_class, stat = "count", fill = country)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = fareblind15) +
  # Labeling
  #---------
  labs(y = "Proportion", x = "CRC Progression Classes") +
  ggtitle("Proportion of Country per CRC Progression Class")

# Location (City)
taxa_meta %>% 
  ggplot(aes(x=disease_class, stat = "count", fill = location)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = fareblind15) +
  # Labeling
  #---------
  labs(y = "Proportion", x = "CRC Progression Classes") +
  ggtitle("Proportion of City per CRC Progression Class")

# Non-Westernized
taxa_meta %>% 
  ggplot(aes(x=disease_class, stat = "count", fill = non_westernized)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = fareblind15) +
  # Labeling
  #---------
  labs(y = "Proportion", x = "CRC Progression Classes") +
  ggtitle("Proportion of 'non-westernized' per CRC Progression Class")

# Disease Location
taxa_meta %>% 
  ggplot(aes(x=disease_class, stat = "count", fill = disease_location)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = fareblind15) +
  # Labeling
  #---------
  labs(y = "Proportion", x = "CRC Progression Classes") +
  ggtitle("Proportion of Disease Location per CRC Progression Class")

# Disease Subtype
taxa_meta %>% 
  ggplot(aes(x=disease_class, stat = "count", fill = disease_subtype)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = fareblind15) +
  # Labeling
  #---------
  labs(y = "Proportion", x = "CRC Progression Classes") +
  ggtitle("Proportion of Disease Subtype per CRC Progression Class")

```

### Technical and Curation-Related Sources of Bias

```{r}
#| label: Technical and Curation-Related
# ---
# Curator
taxa_meta %>% 
  ggplot(aes(x=disease_class, stat = "count", fill = curator)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = fareblind15) +
  # Labeling
  #---------
  labs(y = "Proportion", x = "CRC Progression Classes") +
  ggtitle("Proportion of Curator per CRC Progression Class")

# Sequencing Platform
taxa_meta %>% 
  ggplot(aes(x=disease_class, stat = "count", fill = sequencing_platform)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = fareblind15) +
  # Labeling
  #---------
  labs(y = "Proportion", x = "CRC Progression Classes") +
  ggtitle("Proportion of Sequencing Platform per CRC Progression Class")

# DNA Extraction Kit
taxa_meta %>% 
  ggplot(aes(x=disease_class, stat = "count", fill = DNA_extraction_kit)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = fareblind15) +
  # Labeling
  #---------
  labs(y = "Proportion", x = "CRC Progression Classes") +
  ggtitle("Proportion of DNA Extraction Kit per CRC Progression Class")
```

# Summary Statistics: Metadata

## Disease-Related

```{r}
#| label: Disease Related Summary Statistics
# ---
# Study Condition
table(taxa_meta$study_condition, useNA = "ifany") %>% 
  as.data.frame() %>%          # convert table to dataframe
  kbl(col.names = c("Study Condition", "Count"), 
      caption = "Original Study Condition") %>%
  kable_styling(full_width = FALSE) %>%
  column_spec(2, bold = TRUE)

# Our Proposed CRC Progression Classifiers
table(taxa_meta$disease_class, useNA = "ifany") %>%
  as.data.frame() %>%
  kbl(col.names = c("CRC Progression Class", "Count"), 
      caption = "Proposed CRC Progression Classifier Spread") %>%
  kable_styling(full_width = FALSE) %>%
  column_spec(2, bold = TRUE)

# The diagnosed diseases of patients across the classifiers
table(taxa_meta$disease, taxa_meta$disease_class, useNA = "ifany") %>%
  as.data.frame.matrix() %>%   
  tibble::rownames_to_column("Disease Diagnoses") %>%
  kbl(
    caption = "2-D table of Diagnosed Diseases across CRC Progression Classifiers") %>%
  kable_styling(full_width = FALSE) %>%
  column_spec(1, bold = TRUE)
```

## Age Related

```{r}
#| label: Age Related Summary Statistics
# ---

table(taxa_meta$age_decade, useNA = "ifany") %>%
  as.data.frame() %>%
  kbl(col.names = c("Age Decade", "Count"), 
      caption = "Spread of Age by Decade") %>%
  kable_styling(full_width = FALSE) %>%
  column_spec(2, bold = TRUE)

taxa_meta %>% 
  summarize(
    N = sum(!is.na(age)), 
    Mean = mean(age, na.rm = TRUE), 
    SD = sd(age, na.rm = TRUE),
    Median = median(age, na.rm = TRUE),
    Q1 = quantile(age, 0.25, na.rm = TRUE),
    Q3 = quantile(age, 0.75, na.rm = TRUE)
  )  %>%
  # Make Vertical Table
  # -------------------
  tidyr::pivot_longer(
    everything(),
    names_to = "Statistic",
    values_to = "Value"
  ) %>%
  kbl(digits = 2, caption = "Age Summary Statistics") %>%
  kable_styling(full_width = FALSE) %>%
  column_spec(2, bold = TRUE)

taxa_meta %>% 
  group_by(gender) %>% 
  summarize(
    N = n(), 
    Mean = mean(age, na.rm = TRUE), 
    SD = sd(age, na.rm = TRUE),
    Median = median(age, na.rm = TRUE),
    Q1 = quantile(age, 0.25, na.rm = TRUE),
    Q3 = quantile(age, 0.75, na.rm = TRUE), 
    .groups = "drop"
  )  %>%
  mutate(
    pct = 100 * N / sum(N),
    `N (%)` = sprintf("%d (%.1f%%)", N, pct)
  ) %>%
  select(gender, `N (%)`, Mean, SD, Median, Q1, Q3) %>% 
  kbl(digits = 2, caption = "Age Summary Statistics by Gender") %>%
  kable_styling(full_width = FALSE) %>%
  column_spec(2, bold = TRUE)
```

## Life-Style / Geography Related

### BMI

```{r}
#| label: BMI Related Summary Statistics
# --- 
# BMI
######
taxa_meta %>% 
  summarize(
    N = sum(!is.na(BMI)), 
    Mean = mean(BMI, na.rm = TRUE), 
    SD = sd(BMI, na.rm = TRUE),
    Median = median(BMI, na.rm = TRUE),
    Q1 = quantile(BMI, 0.25, na.rm = TRUE),
    Q3 = quantile(BMI, 0.75, na.rm = TRUE)
  )  %>%
  # Make Vertical Table
  # -------------------
  tidyr::pivot_longer(
    everything(),
    names_to = "Statistic",
    values_to = "Value"
  ) %>%
  kbl(digits = 2, caption = "BMI Summary Statistics") %>%
  kable_styling(full_width = FALSE) %>%
  column_spec(2, bold = TRUE)

taxa_meta %>% 
  group_by(disease_class) %>% 
  summarize(
    N = n(), 
    Mean = mean(BMI, na.rm = TRUE), 
    SD = sd(BMI, na.rm = TRUE),
    Median = median(BMI, na.rm = TRUE),
    Q1 = quantile(BMI, 0.25, na.rm = TRUE),
    Q3 = quantile(BMI, 0.75, na.rm = TRUE), 
    .groups = "drop"
  )  %>%
  select(disease_class, N , Mean, SD, Median, Q1, Q3) %>% 
  kbl(digits = 2, caption = "BMI Summary Statistics by CRC Progression Classifier") %>%
  kable_styling(full_width = FALSE) %>%
  column_spec(2, bold = TRUE)
```

### Geography

```{r}
#| label: Geography Related Summary Statistics
# ---

table(taxa_meta$country, useNA = "ifany") %>%
  as.data.frame() %>%
  arrange(desc(Freq)) %>% 
  kbl(col.names = c("Country", "Count"), 
      caption = "Country Frequency") %>%
  kable_styling(full_width = FALSE) %>%
  column_spec(2, bold = TRUE) 

table(taxa_meta$location, useNA = "ifany") %>%
  as.data.frame() %>%
  arrange(desc(Freq)) %>% 
  kbl(col.names = c("City", "Count"), 
      caption = "City Frequency") %>%
  kable_styling(full_width = FALSE) %>%
  column_spec(2, bold = TRUE)
```

# PCoA

## Assumptions
