---
title: "Metagenomic EDA of Fecal Samples"
format: pdf
author:
    - name: Kaitlyn Schisler
    - name: Cory Spern
    - name: Faith Shipman
    - name: Jacob Anderson
---

# Delete Before Submitting

+--------------+---------------------------------------------------------------+
| Student      | Assignment                                                    |
+==============+===============================================================+
| Cory         | Country & Category                                            |
+--------------+---------------------------------------------------------------+
| Kaitlyn      | Age & Sex                                                     |
|              |                                                               |
|              | Data Availability Table (Show's what cMD has from each study) |
+--------------+---------------------------------------------------------------+
| Faith        | Limitations (Missing Data, Pop size, etc.)                    |
+--------------+---------------------------------------------------------------+
| Jacob        | Summary Stats, Glossary & Introduction                        |
+--------------+---------------------------------------------------------------+

### Assignment Guidelines from Canvas

[**Step 1: EDA Objectives & Preparing data**]{.underline}

-   define at least 3 EDA objectives to investigate (necessary variables? distributions conducive to modelling? data quality issues? is data suitable? representative of population?)

-   There's a document available about what's good to do for EDA on canvas that we can look at if needed

[**Step 2: Conduct EDA**]{.underline}

-   Plots/visualizations, summary stats, tables

-   document findings clearly

-   Code, additional plots & tables that aren't the main talking points can be included in an appendix section.

[**Step 3: EDA Report (Single spaced, 12 pt. font)**]{.underline}

-   **Introduction to Data (1-2 page)**\
    Provide an overview of your dataset, including its source (with relevant links), purpose, and your rationale for selecting it. Clearly state your EDA objectives and briefly describe the EDA methods used. Explain the importance of EDA in the context of your overall project.

    -   Pull from methods section from proposal/revised proposal and the project statement, good depth there on selection of the data and why we chose the data/methods.

-   **Exploratory Data Analysis (4-5 pages)**\
    This is the core of your report. Present your findings using relevant summary statistics, tables, and visualizations. Ensure each EDA objective is addressed clearly and supported by appropriate evidence from your analysis.

    -   Flow of this section will be important. We can play with the flow once everyone's section is complete.

-   **Summary (1-2 pages)**\
    Summarize your key findings and highlight the main takeaways from your EDA. Describe the next steps for your project, including how the insights gained from EDA will inform your choice of data modeling techniques for the main analysis. Briefly outline which techniques you plan to implement next and why.

### General Notes

-   Potential flow of our EDA section: Data Availability & Selection → Metadata Analysis → Taxa & Pathway Analysis → Data Limitations

    -   Data Availability

        -   What does the whole cMD3 offer in its standardized data, including how Bioconductor data works (`summarizedExperiments` vs `TreeSummarizedExperiments`)?

        -   What data remains, what could be answered, what can still be explored?

        -   ***Our original plan was to include metastases, which our data set has none so that is no longer possible***, so that could be mentioned here in this EDA section.

    -   Metadata EDA

        -   The factors we mentioned in our proposals (ie age, sex, geography/ethnicity, diet, antibiotics, etc)

            -   What has too little to give real meaning behind trends we see?

        -   For geography, the top 3 affected countries by CRC are represented in our data set, which is good. We should note that we are missing representations from certain continents

    -   Taxa vs pathways

        -   what we can do now (personal machines) vs what we need to do later on the HPC cluster (Sol)

    -   Data Limitations

        -   What gets lost once we filter it with our exclusionary criteria?

        -   What would be unable to answer with what we have, and what gaps would remain to be filled?

------------------------------------------------------------------------

------------------------------------------------------------------------

```{r, echo=FALSE}
#| label: Setup
source("Qualifying_Studies_List.R")
suppressPackageStartupMessages({
  library(dplyr)
  library(knitr)
  library(kableExtra)
  library(mosaic)
  library(ggplot2)
  library(grid)
  library(gridExtra)
  library(knitr)
  library(patchwork)
  library(tidyr)
})

progression_order <- c(
  "Other", 
  "HC",
  "PA",
  "PA+",
  "PA-M",
  "CRC",
  "CRC+",
  "CRC-H",
  "CRC-M"
)

progression_colors <- c(
  "HC" = "#539deb",
  "PA" = "#e4e85b",
  "PA+" = "#e8a25b",
  "CRC" = "#cf1919",
  "CRC+" = "#8f0000",
  "CRC-M" = "#4d0404",
  "CRC-H" = "#520f76",
  "Other" = "#aaa3a3"
)

qualifying_studies <- qualifying_studies %>% 
  mutate(
    disease_class = factor(disease_class, levels = progression_order)
  )
```

### Introduction

### Glossary

Yada Yada Yada

##### Diagnostic Categories

+--------------+-------------------------------------------------------------------------+
| Abbreviation | Description                                                             |
+==============+=========================================================================+
| HC           | Healthy control                                                         |
+--------------+-------------------------------------------------------------------------+
| PA           | Polyp / adenoma                                                         |
+--------------+-------------------------------------------------------------------------+
| PA+          | Polyp / adenoma with comorbidities                                      |
+--------------+-------------------------------------------------------------------------+
| CRC          | Colorectal cancer                                                       |
+--------------+-------------------------------------------------------------------------+
| CRC-H        | History of colorectal cancer with resection                             |
+--------------+-------------------------------------------------------------------------+
| CRC+         | Colorectal cancer with comorbidities                                    |
+--------------+-------------------------------------------------------------------------+
| Other        | Any diagnosed disease without current or history of CRC-related growths |
+--------------+-------------------------------------------------------------------------+

### Exploratory Data Analysis

#### Data Availability

#### Metadata Analysis (Kaitlyn)

Of great benefit to having the cMD3 database is the standardization of the metadata, allowing for the analysis of potential confounding variables for biases and major discrepancies in the data that may influence the results and interpretations downstream. With prior evidence supporting age, gender, diet, geographical location, BMI, and drinking and smoking habits are all factors that may influence the relative risk of CRC (SOURCE), they are important variables to consider in our initial EDA, checking for bias in distribution and sources of significance.

##### Normality and Summary Statistics

All samples included contain information regarding the age, which is important considering it being one of the factors in CRC screening and preventative care (SOURCE). Summary statistics were conducted (Appendix 1-A), showing a mean age of 59, standard deviation of 14.75, and median of 63, suggesting a skew in this continuous variable. When faceted by gender (Appendix 1-B), aside from the proportion of males and females being statistically different from each other (41.1% female, **CALCULATE P VALUE?**), the median, mean, and quartiles are not remarkably different. A quantile-quantile (QQ) plot (Figure 1-A) and histogram (Figure 1-B) were generated to corroborate this information, checking for extreme skewedness. These show that while there is a slight left tailed skew in age with a bimodal pattern , it is not to an extreme that would harm our final model detrimentally. Random Forest models are robust to skew (SOURCE). Skew and variance of age was also checked between the sexes via a faceted histogram, which can be found in the supplementary visualizations (Appendix 2-A).

```{r, echo=FALSE}
#| label: Normality Check Age Plain
# ---

# QQPlot
########
normality1 <- qualifying_studies %>% 
  # Set Up
  # ------
  filter(!is.na(age)) %>%
  # Plotting
  # --------
  ggplot(aes(sample = age)) + 
  geom_qq(size = 1, alpha = 0.8) + 
  geom_qq_line() +
  # Labeling
  #---------
  labs(title = "QQ Plot of Age") +
  theme(plot.title = element_text(hjust = 0, size = 10), 
        axis.title = element_text(size = 10))

# Histogram
###########
normality2 <- qualifying_studies %>%
  # Set Up
  # ------
  filter(!is.na(age)) %>%
  # Plotting
  # --------
  ggplot(aes(x = age)) + 
  geom_histogram(bins=20) + 
  # Labeling
  #---------
  labs(title = "Distribution of Age", x = "Age", y = "Count") +
  theme(plot.title = element_text(hjust = 0, size = 10), 
        axis.title = element_text(size = 10))

# Display
#########

normality_caption <- str_wrap(
  "(A) QQ plot showing slight deviation from the normality line, indicating mild skew. 
   (B) Histogram demonstrating a bimodal age distribution with a secondary peak near 25 years and a primary peak near 60 years.",
  width = 95 # maximum width
)

(normality1 | normality2) +
  plot_annotation(
    title = "Figure 1: Overall Age Normality Check",
    tag_levels = "A",
    caption = normality_caption
  ) +
  theme(
    plot.title = element_text(hjust = 0, size = 10),
    plot.caption = element_text(hjust = 0, size = 8)
  )
```

```{r, echo=FALSE}
#| label: CRC Progression Classifier Spread
# ---

# Barplot of age categories: Colored
####################################
decade1 <- qualifying_studies %>% 
  # Set Up 
  # ------ 
  filter (!is.na(age_decade)) %>% 
  # Plotting 
  # -------- 
  ggplot(aes(x=age_decade, fill = disease_class)) + 
  geom_bar() + 
  scale_fill_manual(values = progression_colors, 
                    name = "CRC Progression \nClassifiers") + 
  # Labeling 
  #--------- 
  theme(axis.text.x = element_text(angle = 30, hjust = 1, lineheight = 1.1), 
        axis.title.x = element_text(hjust = 0.5), 
        plot.title = element_text(size = 11),
        legend.position = "none") + 
  labs(
    title = "Age Distribution By Decade", 
    x = "Age by Decade", y = "Count")

# Proportion: Age Decade
###########################
decade2 <- qualifying_studies %>%
  # Set Up
  # ------
  filter (!is.na(age_decade)) %>%
  # Plotting
  # --------
  ggplot(aes(x=age_decade, fill = disease_class)) +
  geom_bar(position="fill") +
  scale_fill_manual(values = progression_colors, 
                    name = "CRC Progression \nClassifiers") +
  # Labeling
  #---------
  labs(title = "Proportion of CRC Progression by Age Decade",
       x = "Age By Decade", y = "Proportion") +
  theme(axis.text.x = element_text(
    angle = 30, hjust = 1, lineheight = 1.1), 
    legend.position = "none", 
    plot.title = element_text(size = 11)
  )

# Gendered Barplot: Proportions
###############################
decade3 <- qualifying_studies %>%
  # Set Up
  # ------
  filter(!is.na(age_decade), !is.na(gender)) %>%
  # Plotting
  # --------
  ggplot(aes(x = age_decade, fill = disease_class)) + 
  geom_bar(position = "fill") +
  scale_fill_manual(values = progression_colors, 
                    name = "CRC Progression \nClassifiers") +
  facet_wrap(~gender) +
  # Labeling
  #---------
  labs(title = "Proportion of CRC Progression by Age Decade, \nGrouped by Gender", 
       x = "Age by Decade", y = "Proportion") +
  theme(axis.text.x = element_text(angle = 25, hjust = 1, lineheight = 1.1), 
        legend.position = "none", 
        plot.title = element_text(size = 12))

# Age by Decade Patchwork
#########################
layout1 <- c( # t, l , b = t, r = l
  area(1, 1, 2, 2), 
  area(1, 3, 2, 4), 
  area(3, 1, 4, 4)
)

#plot(layout1)

decade1 + decade2 + decade3 + plot_layout(design = layout1, guides = "collect") +
  plot_annotation(
    title = "Figure 2: CRC Progression Classification Variability",
    tag_levels = "A",
    caption = "Placeholder Caption"
  ) +
  theme(legend.position = "right", 
    plot.title = element_text(hjust = 0, size = 10),
    plot.caption = element_text(hjust = 0, size = 8), 
    legend.key.size = unit(0.25, "cm")
  )
```

#### Taxa & Pathway Analysis (Cory)

#### Data Limitations

### Conclusion

### References
