---
title: "Supplemental Materials"
subtitle: "Metagenomic Analysis of Fecal Samples as a Differential Diagnostic Tool for Colorectal Cancer"

author: 
  - name: Kaitlyn Schisler
  - name: Cory Spern
  - name: Faith Shipman
  - name: Jacob Anderson

format: 
  pdf:
    toc: true
    toc-depth: 3
    geometry: 
      - margin=0.5in
    header-includes:
      - \setkomafont{title}{\Large\bfseries}
      - \setkomafont{author}{\normalsize}
      - \setkomafont{date}{\small}
---

```{r}
#| label: Setup
#| echo: false
#| message: false
suppressPackageStartupMessages({
  library(tidyverse)
  library(knitr)
  library(kableExtra)
  library(mosaic)
  library(grid)
  library(gridExtra)
  library(patchwork)
  library(curatedMetagenomicData)
})

# Run and load any data files needed prior to running this file, since we are no longer using Qualifying_Studies_List.R
eda_meta_df <- readRDS("Kaitlyn_EDA/data/eda_meta_df.rds")
# from kaitlyn's Merged_Studies_EDA.R
progression_order <- c(
  "HC",
  "PA",
  "PA+",
  "PA-M",
  "CRC",
  "CRC+",
  "CRC-H",
  "CRC-M", 
  "Other"
)

progression_colors <- c(
  "HC" = "#539deb",
  "PA" = "#e4e85b",
  "PA+" = "#e8a25b",
  "PA-M" = "#d4700d",
  "CRC" = "#cf1919",
  "CRC+" = "#8f0000",
  "CRC-M" = "#4d0404",
  "CRC-H" = "#520f76",
  "Other" = "#aaa3a3"
)

progression_legend_title <- "CRC Progression\nClassifiers"
```

## 1 - Summary Tables

### 1-A: Age Summary Statistics

```{r}
#| label: Overall Age Summary Statistics
#| echo: false
#| message: false

# Stats Setup
#############
age_overall <- eda_meta_df %>% 
  # Summary Stats
  # -------------
  summarize(
    n = sum(!is.na(age)), 
    mean = mean(age, na.rm = TRUE), 
    sd = sd(age, na.rm = TRUE),
    median = median(age, na.rm = TRUE),
    Q1 = quantile(age, 0.25, na.rm = TRUE),
    Q3 = quantile(age, 0.75, na.rm = TRUE)
  )  %>%
  # Make Vertical Table
  # -------------------
  pivot_longer(
    everything(),
    names_to = "Statistic",
    values_to = "Value"
  ) %>% 
  mutate(Statistic = recode(Statistic,
  n = "N",
  mean = "Mean",
  sd = "Standard Deviation",
  median = "Median",
  Q1 = "First Quartile (Q1)",
  Q3 = "Third Quartile (Q3)"
))

# Kable Table
#############
age_overall %>% 
  kbl(
    col.names = c("Statistic", "Value"),
    caption = "Overall Age Summary",
    digits = 2,
    format = "latex"
  ) %>%
  kable_styling(full_width = FALSE, latex_options = "HOLD_position") %>%
  column_spec(1, bold = TRUE)
```

### 1-B: Age by Gender Summary Statistics

```{r}
#| label: Age by Gender
#| echo: false
#| message: false
# ---

# Stats Setup
#############
age_x_gender <- eda_meta_df %>% 
  group_by(gender) %>% 
  summarize(
    n = n(), 
    mean = mean(age, na.rm = TRUE), 
    sd = sd(age, na.rm = TRUE),
    median = median(age, na.rm = TRUE),
    Q1 = quantile(age, 0.25, na.rm = TRUE),
    Q3 = quantile(age, 0.75, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  mutate(
    pct = 100 * n / sum(n),
    `N (%)` = sprintf("%d (%.1f%%)", n, pct)
  ) 

# Table Setup
#############
age_x_gender %>% 
  select(gender, `N (%)`, mean, sd, median, Q1, Q3) %>%
  kbl(
    col.names = c("Gender", "N (%)", "Mean", "Standard Deviation",
                  "Median", "Q1", "Q3"),
      caption = "Age Summary by Gender", format = "latex", digits = 2) %>%
  kable_styling(full_width = F, latex_options = "hold_position") %>% 
  column_spec(1, bold = T)

```

### 1-C: Age by Decade Summary Statistics

```{r}
#| label: Age by Decade Summary
#| echo: false
#| message: false
# Stats Setup
#############
age_decade <- eda_meta_df %>% 
  # Summary Stats
  # -------------
  group_by(age_decade) %>% 
  summarize(
  n = n(), 
  .groups = "drop") %>% 
  mutate(
    pct = 100 * n / sum(n),
    `N (%)` = sprintf("%d (%.1f%%)", n, pct)
  )

# Kable Table
#############
age_decade %>% 
  select(age_decade, `N (%)`) %>%
  kbl(
    col.names = c("Age (Decade)", "N (%)"),
      caption = "Age Summary by Decade", format = "latex", digits = 2) %>%
  kable_styling(full_width = F, latex_options = "hold_position") %>% 
  column_spec(1, bold = T)
```

### 1-D: Country Counts and Proportions

```{r}
#| label: Country Counts and Proportions
#| echo: false
#| message: false

# Count and Proportion
######################
country_overall <- eda_meta_df %>% 
  # Summary Stats
  # -------------
  group_by(country) %>% 
  summarize(
  n = n(), 
  .groups = "drop") %>% 
  mutate(
    pct = 100 * n / sum(n),
    `N (%)` = sprintf("%d (%.1f%%)", n, pct)
  ) %>% 
  arrange('Country')

# Kable Table
#############
country_overall %>% 
  select(country, `N (%)`) %>%
  kbl(
    col.names = c("Country", "N (%)"),
      caption = "Country Overall Summary (NATO country code, trigram)", 
    format = "latex", digits = 2) %>%
  kable_styling(full_width = F, latex_options = "HOLD_position") %>% 
  column_spec(1, bold = T)
```


## 2 - Supplemental Visualizations

### 2-A: Age QQ-Plot (Supports Histogram)

```{r}
#| label: Age Normality Check - QQ Plot
#| echo: false
#| message: false

# QQPlot
########
eda_meta_df %>% 
  # Set Up
  # ------
filter(!is.na(age)) %>%
  # Plotting
  # --------
ggplot(aes(sample = age)) + 
  geom_qq(size = 1, alpha = 0.8) + 
  geom_qq_line() +
  # Labeling
  #---------
labs(title = "QQ Plot of Age") +
  theme(plot.title = element_text(hjust = 0, size = 10), 
        axis.title = element_text(size = 10))
```

### 2-B: Age Histogram by Gender

```{r}
#| label: Age Normality Check - Gendered
#| echo: false
#| message: false
# Gendered Histogram
####################
eda_meta_df %>%
  # Set Up
  # ------
  filter(!is.na(age), !is.na(gender)) %>%
  # Plotting
  # --------
  ggplot(aes(x = age)) + 
  geom_histogram(bins = 30, fill = "#539deb", color = "black") +
  facet_wrap(~gender) +
  # Labeling
  #---------
  labs(title = "Distribution of Age by Gender", x = "Age", y = "Count",) +
  theme(plot.title = element_text(hjust = 0), 
        axis.title = element_text(size = 10))
```

### 2-C: Country Distribution and Proportions

```{r}
#| label: Countries Associations
#| fig-width: 14
#| fig-height: 6
#| fig-dpi: 500
#| echo: false
#| message: false
# ---

# Countries ~~~~~~~~~~~~~~~~
############################

# Country Count
###############
country1 <- eda_meta_df  %>% 
  # Set Up
  # ------
  filter(!is.na(country)) %>%
  # Plotting
  # --------
  ggplot(aes(x = reorder(country, country, 
                         FUN = function(x) -length(x)))) + 
  geom_bar() +
  # Labeling
  #---------
  geom_text(stat = "count", aes(label = after_stat(count)), 
            vjust = -0.21, size = 3, color = "black") +
  labs(
    title = "Country Sample Count Distribution", 
    x = "Country (NATO country code, trigram)", 
    y = "Count") +
  theme( 
        legend.position = "none", 
        plot.title = element_text(size = 12))

# Country Fill
##############
country2 <- eda_meta_df  %>% 
  # Set Up
  # ------
  filter(!is.na(country)) %>%
  # Plotting
  # --------
  ggplot(aes(x = reorder(country, country, 
                         FUN = function(x) -length(x)), 
             fill=disease_class)) + 
  geom_bar() +
  scale_fill_manual(values = progression_colors, 
                    name = "CRC Progression Classifiers") +
  # Labeling
  #---------
  labs(title = "Country Sample Count Distribution",
       subtitle = "with CRC Progression Classifiers",
       x = "Country (NATO country code, trigram)", 
       y = "Count") +
  theme( 
        legend.position = "none", 
        plot.title = element_text(size = 12))

# Countries Proportions
#######################
country3 <- eda_meta_df %>% 
  # Set Up
  # ------
  filter(!is.na(country)) %>%
  # Plotting
  # --------
  ggplot(aes(x = reorder(country, country, 
                         FUN = function(x) -length(x)), fill = disease_class)) + 
  geom_bar(position = "fill") +
  scale_fill_manual(values = progression_colors, 
                    name = "CRC Progression Classifiers") +
  # Labeling
  #---------
  labs(
    title = "Proportion of CRC Progression Per Country", 
    x = "Country (NATO country code, trigram)", 
    y = "Proportion") +
  theme( 
        plot.title = element_text(size = 12))


# Display
#########
layout3 <- c( # t, l , b = t, r = l
  area(1, 1, 2, 2),
  area(1, 3, 2, 4),
  area(1, 5, 2, 6)
)

#plot(layout3)

geography_caption <- str_wrap(
  "(A) Bar plot of country sample count, sorted in descending order. 
   (B) Stacked bar plot of country sample count, sorted in descending order, filled with CRC progression classifiers.
   (C) Stacked bar plot of the CRC progression classifier proportion for each country in the multi-cohort study.", 
  width = 200
)

country1 + country2 + country3 + 
  plot_layout(design = layout3, guides = "collect", heights = c(1, 0.18)) +
  plot_annotation(
    tag_levels = "A",
    caption = geography_caption
  ) &
  theme(
        plot.title = element_text(hjust = 0, size = 14),
        plot.caption = element_text(hjust = 0, size = 11), 
        legend.key.size = unit(0.5, "cm"),
        legend.title = element_text(size = 10), 
        legend.direction = "vertical", 
        legend.position = "right"
  )
```

### 2-D: CRC Progression Classifier Counts

```{r}
#| label: CRC Progression Classifier Counts
#| echo: false
#| message: false

# Bar Chart of Total CRC Progression Class Proportions
#########################################################
class_count <- eda_meta_df %>%
  # Plotting
  # --------
  ggplot(aes(x=disease_class, stat = "count", fill = disease_class)) +
  geom_bar() +
  scale_fill_manual(values = progression_colors, 
                    name = "CRC Progression \nClassifiers") +
  theme(axis.text.x = element_text(
    angle = 20, hjust = 1, lineheight = 1.1, size = 8), 
    plot.title = element_text(size = 10))+
  # Labeling
  #---------
  labs(y = "Sample Count", x = "CRC Progression Classes") +
  ggtitle("Counts of CRC Progression Classes in Study")

class_count
```

## 3 - Statistical Analyses

### 3-A: CRC Progression Classifier and Age ANOVA Significant Tukey's HSD Pairings

| Disease Pairing | 95% CI Lower | 95% CI Upper | P Value   |
| --------------- | ------------ | ------------ | --------- |
| HC-Other        | -10.50284389 | -1.7236952   | 0.0005460 |
| PA+-HC          |   1.52805911 | 12.0253580   | 0.0020679 |
| CRC-HC          |   1.26185019 |  5.2375002   | 0.0000151 |
| CRC+-HC         |   3.25756254 | 11.8241879   | 0.0000019 |
| CRC+-CRC        |   0.01085945 |  8.5715406   | 0.0488451 |
| CRC-H-CRC+      | -14.81785403 | -1.1021460   | 0.0097263 |

```{r}
#| label: Age vs CRC Progression ANOVA
#| eval: false
#| message: false
#| include: false

age_CRC_ANOVA <- eda_meta_df %>%
  filter(!is.na(age)) 

age_CRC_ANOVA2 <- aov(age ~ disease_class, data = age_CRC_ANOVA)
  
summary(age_CRC_ANOVA2)

TukeyHSD(age_CRC_ANOVA2, conf.level=.95)
```

### 3-B: CRC Progression Classifier and BMI ANOVA Significant Tukey's HSD Pairings

| Disease Pairing | 95% CI Lower | 95% CI Upper | P Value   |
| --------------- | ------------ | ------------ | --------- |
| HC-Other        | -4.94428952  | -1.80320222  | 0.0000000 |
| PA-Other        | -4.13977278  | -0.57390021  | 0.0013977 |
| CRC-Other       | -4.53643961  | -1.39667564  | 0.0000002 |
| CRC-H-Other     | -7.40515642  | -2.45205913  | 0.0000000 |
| PA+-HC          |  2.37307016  |  6.12841806  | 0.0000000 |
| CRC+-HC         |  0.73316259  |  3.81667022  | 0.0001705 |
| PA+-PA          |  1.17521532  |  5.29245416  | 0.0000411 |
| CRC-H-PA        | -4.78350442  | -0.36003813  | 0.0095047 |
| CRC-PA+         | -5.72067642  | -1.96643531  | 0.0000000 |
| CRC-H-PA+       | -8.48746084  | -3.12375118  | 0.0000000 |
| CRC+-CRC        |  0.32664837  |  3.40880795  | 0.0054141 |
| CRC-H-CRC+      | -6.28817020  | -1.37138642  | 0.0000502 |

```{r}
#| label: BMI vs CRC Progression ANOVA
#| eval: false
#| message: false
#| include: false

bmi_CRC_ANOVA <- eda_meta_df %>%
  filter(!is.na(BMI)) 

bmi_CRC_ANOVA2 <- aov(BMI ~ disease_class, data = bmi_CRC_ANOVA)
  
summary(bmi_CRC_ANOVA2)

TukeyHSD(bmi_CRC_ANOVA2, conf.level=.95)
```

### 3-D: Linear Regression Model by CRC Progression Classifier Significant Results

| Coefficients        | Estimate | P Value  |
| ------------------- | -------- | -------- |
| Other               | 2.793178 | 7.00e-07 |
| PA                  | 1.012194 | 0.004601 |
| PA+                 | 4.852519 | 4.19e-12 |
| CRC+                | 2.008241 | 0.000225 |
| Age centered: Other | 0.158122 | 0.017010 |
| Age centered: CRC-H | 0.135257 | 0.012499 |

```{r}
#| label: LM to match faceted scatterplot
#| eval: false
#| message: false
#| include: false

eda_meta_df$age_centered <- scale(
  eda_meta_df$age,
  center = TRUE,
  scale = FALSE
)

eda_meta_df$disease_class <- relevel(
  eda_meta_df$disease_class,
  ref = "HC" # Changed to refer to healthy control rather than "Other"
)

age_bmi_scatterplot_lm <- lm(BMI ~ age_centered * disease_class, 
               data = eda_meta_df)

summary(age_bmi_scatterplot_lm)

```

### 3-E: Examples of Attempted RF Models for Hyperparameter Tuning

| Packages       | Parameters                                                           | Classifiers                    | Predictors                                                                                 | Preprocessing                                                                                                                                            | Accuracy                                                   | Confusion Matrix Results                                                                                        | Per -Class F1 Scores                                     | Per-Class AUC-ROC                                                                     |
| -------------- | -------------------------------------------------------------------- | ------------------------------ | ------------------------------------------------------------------------------------------ | -------------------------------------------------------------------------------------------------------------------------------------------------------- | ---------------------------------------------------------- | --------------------------------------------------------------------------------------------------------------- | -------------------------------------------------------- | ------------------------------------------------------------------------------------- |
| ranger & caret | Trees: 500                                                           | HC PA CRC  <br>Other PA+ CRC + | Taxa only                                                                                  | 70/30 study aware manual split<br>0.001% prevalence threshold on TRAIN, applied to all data<br>Remove near zero variance from TRAIN, applied to all data | Out-Of-Bag(OOB): 0.356                                     | Accuracy: 0.4916  <br>P-value: 9.88e-12                                                                         | CRC: 0.61  <br>HC: 0.57  <br>All other classes showed NA | CRC 0.758  <br>CRC+ 0.625  <br>HC 0.709  <br>Other 0.517  <br>PA 0.567  <br>PA+ 0.531 |
| ranger & caret | Trees: 500<br>                                                       | HC PA CRC  <br>Other           | Taxa only                                                                                  | 70/30 study aware manual split<br>0.001% prevalence threshold on TRAIN, applied to all data<br>Remove near zero variance from TRAIN, applied to all data | OOB: 0.298                                                 | Accuracy: 0.5849  <br>P-value: 2.2e-16                                                                          | CRC: 0.70  <br>HC: 0.64  <br>PA: 0.03  <br>Other: NA     | CRC 0.843  <br>HC 0.757  <br>Other 0.586  <br>PA 0.536                                |
| ranger & caret | Trees: 500<br>mtry: 3<br>Target Node: 5                              | HC PA CRC                      | Taxa only                                                                                  | 70/30 study aware manual split<br>0.001% prevalence threshold on TRAIN, applied to all data<br>Remove near zero variance from TRAIN, applied to all data | OOB: 0.263                                                 | Accuracy: 0.568  <br>P-value: 5.938e-11                                                                         | CRC: 0.636  <br>HC: 0.631  <br>PA: NA                    | CRC 0.741  <br>HC 0.689  <br>PA 0.547                                                 |
| ranger & caret | Trees: 500<br>mtry: 3<br>Target Node: 5                              | HC PA CRC                      | Taxa only                                                                                  | LOSO split<br>0.001% prevalence threshold on TRAIN, applied to all data<br>Remove near zero variance from TRAIN, applied to all data<br>                 | Mean LOSO accuracy: 0.651  <br>Median LOSO accuracy: 0.678 | Individual plots generated rather than a data frame for averaging. This will need to be revised for future runs | Medians  <br>CRC: 0.706  <br>HC: 0.708  <br>PA: NA       | Medians  <br>CRC: 0.809  <br>HC: 0.763  <br>PA: 0.562                                 |
| ranger & caret | Trees: 500<br>Target Node: 5<br>metadata completeness threshold: 90% | HC PA CRC                      | Taxa, Pathway Cov, Pathway Abundance, Metadata columns that satisfy the criteria in params | 70/30 study aware manual split<br>0.001% prevalence threshold on TRAIN, applied to all data<br>Remove near zero variance from TRAIN, applied to all data | OOB: 0.276                                                 | Accuracy: 0.5473  <br>P-value: 1.17e-8                                                                          | CRC: 0.602  <br>HC: 0.618  <br>PA: NA                    | CRC: 0.715  <br>HC: 0.677  <br>PA: 0.552                                              |

## 4 - Relevant Links

### 4-A: GitHub for Script Availability

Link to our project GitHub Repository: <https://github.com/chejj/Therayess>

### 4-B: cMD3 Data Set

Link to the cMD3 data set Bioconductor Page: <https://bioconductor.org/packages/3.16/data/experiment/html/curatedMetagenomicData.html>
