---
title: "Supplemental Materials"
subtitle: "Metagenomic Analysis of Fecal Samples as a Differential Diagnostic Tool for Colorectal Cancer"

author: 
  - name: Kaitlyn Schisler
  - name: Cory Spern
  - name: Faith Shipman
  - name: Jacob Anderson

format: 
  pdf:
    toc: true
    toc-depth: 3
    header-includes:
      - \setkomafont{title}{\Large\bfseries}
      - \setkomafont{author}{\normalsize}
      - \setkomafont{date}{\small}
---

```{r}
#| label: Setup
#| echo: false
#| message: false
suppressPackageStartupMessages({
  library(tidyverse)
  library(knitr)
  library(kableExtra)
  library(mosaic)
  library(grid)
  library(gridExtra)
  library(patchwork)
  library(curatedMetagenomicData)
})

# Run and load any data files needed prior to running this file, since we are no longer using Qualifying_Studies_List.R
merged_studies <- readRDS("Kaitlyn_EDA/data/merged_studies.rds")
eda_meta_df <- as.data.frame(colData(merged_studies[["relative_abundance"]]))
# from kaitlyn's Merged_Studies_EDA.R
progression_order <- c(
  "HC",
  "PA",
  "PA+",
  "PA-M",
  "CRC",
  "CRC+",
  "CRC-H",
  "CRC-M", 
  "Other"
)

progression_colors <- c(
  "HC" = "#539deb",
  "PA" = "#e4e85b",
  "PA+" = "#e8a25b",
  "PA-M" = "#d4700d",
  "CRC" = "#cf1919",
  "CRC+" = "#8f0000",
  "CRC-M" = "#4d0404",
  "CRC-H" = "#520f76",
  "Other" = "#aaa3a3"
)

progression_legend_title <- "CRC Progression\nClassifiers"
```

## 1 - Summary Tables

### 1-A: Age Summary Statistics

```{r}
#| label: Overall Age Summary Statistics
#| echo: false
#| message: false
# Stats Setup
#############
age_overall <- eda_meta_df %>% 
  # Summary Stats
  # -------------
  summarize(
    n = sum(!is.na(age)), 
    mean = mean(age, na.rm = TRUE), 
    sd = sd(age, na.rm = TRUE),
    median = median(age, na.rm = TRUE),
    Q1 = quantile(age, 0.25, na.rm = TRUE),
    Q3 = quantile(age, 0.75, na.rm = TRUE)
  )  %>%
  # Make Vertical Table
  # -------------------
  pivot_longer(
    everything(),
    names_to = "Statistic",
    values_to = "Value"
  ) %>% 
  mutate(Statistic = recode(Statistic,
  n = "N",
  mean = "Mean",
  sd = "Standard Deviation",
  median = "Median",
  Q1 = "First Quartile (Q1)",
  Q3 = "Third Quartile (Q3)"
))

# Kable Table
#############
age_overall %>% 
  kbl(
    col.names = c("Statistic", "Value"),
    caption = "Overall Age Summary",
    digits = 2,
    format = "latex"
  ) %>%
  kable_styling(full_width = FALSE) %>%
  column_spec(1, bold = TRUE)
```

### 1-B: Age by Gender Summary Statistics

```{r}
#| label: Age by Gender
#| echo: false
#| message: false
# ---

# Stats Setup
#############
age_x_gender <- eda_meta_df %>% 
  group_by(gender) %>% 
  summarize(
    n = n(), 
    mean = mean(age, na.rm = TRUE), 
    sd = sd(age, na.rm = TRUE),
    median = median(age, na.rm = TRUE),
    Q1 = quantile(age, 0.25, na.rm = TRUE),
    Q3 = quantile(age, 0.75, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  mutate(
    pct = 100 * n / sum(n),
    `N (%)` = sprintf("%d (%.1f%%)", n, pct)
  ) 

# Table Setup
#############
age_x_gender %>% 
  select(gender, `N (%)`, mean, sd, median, Q1, Q3) %>%
  kbl(
    col.names = c("Gender", "N (%)", "Mean", "Standard Deviation",
                  "Median", "Q1", "Q3"),
      caption = "Age Summary by Gender", format = "latex", digits = 2) %>%
  kable_styling(full_width = F) %>% 
  column_spec(1, bold = T)

```

### 1-C: Age by Decade Summary Statistics

```{r}
#| label: Age by Decade Summary
#| echo: false
#| message: false
# Stats Setup
#############
age_decade <- eda_meta_df %>% 
  # Summary Stats
  # -------------
  group_by(age_decade) %>% 
  summarize(
  n = n(), 
  .groups = "drop") %>% 
  mutate(
    pct = 100 * n / sum(n),
    `N (%)` = sprintf("%d (%.1f%%)", n, pct)
  )

# Kable Table
#############
age_decade %>% 
  select(age_decade, `N (%)`) %>%
  kbl(
    col.names = c("Age (Decade)", "N (%)"),
      caption = "Age Summary by Decade", format = "latex", digits = 2) %>%
  kable_styling(full_width = F) %>% 
  column_spec(1, bold = T)
```

### 1-D: Country Summary Statistics

```{r}
#| label: Country Summary Statistics
#| echo: false
#| message: false

# Count and Proportion
######################
country_overall <- eda_meta_df %>% 
  # Summary Stats
  # -------------
  group_by(country) %>% 
  summarize(
  n = n(), 
  .groups = "drop") %>% 
  mutate(
    pct = 100 * n / sum(n),
    `N (%)` = sprintf("%d (%.1f%%)", n, pct)
  ) %>% 
  arrange('Country')

# Kable Table
#############
country_overall %>% 
  select(country, `N (%)`) %>%
  kbl(
    col.names = c("Country", "N (%)"),
      caption = "Country Overall Summary (NATO country code, trigram)", 
    format = "latex", digits = 2) %>%
  kable_styling(full_width = F) %>% 
  column_spec(1, bold = T)

# Country Classifier Count
##########################
country_overall <- eda_meta_df %>% 
  filter(!is.na(country)) %>%
  group_by(country, disease_class) %>% 
  summarise(n = n(), .groups = "drop") %>% 
  arrange(country, desc(n))

# Classifier Kable Table
########################
country_overall %>% 
  kbl(col.names = c("Country", "CRC Classifier", "Count"),
      caption = "Country CRC Progression Summary (NATO country code, trigram)", 
      format = "latex") %>%
  kable_styling(full_width = F) %>% 
  column_spec(1, bold = T) %>% 
  collapse_rows(columns = 1, valign = "top")

```

## 2 - Supplemental Visualizations

### 2-A: Age QQ-Plot (Supports Histogram)

```{r}
#| label: Age Normality Check - QQ Plot
#| echo: false
#| message: false

# QQPlot
########
eda_meta_df %>% 
  # Set Up
  # ------
filter(!is.na(age)) %>%
  # Plotting
  # --------
ggplot(aes(sample = age)) + 
  geom_qq(size = 1, alpha = 0.8) + 
  geom_qq_line() +
  # Labeling
  #---------
labs(title = "QQ Plot of Age") +
  theme(plot.title = element_text(hjust = 0, size = 10), 
        axis.title = element_text(size = 10))
```

### 2-B: Age Histogram by Gender

```{r}
#| label: Age Normality Check - Gendered

# Gendered Histogram
####################
eda_meta_df %>%
  # Set Up
  # ------
  filter(!is.na(age), !is.na(gender)) %>%
  # Plotting
  # --------
  ggplot(aes(x = age)) + 
  geom_histogram(bins = 30, fill = "#539deb", color = "black") +
  facet_wrap(~gender) +
  # Labeling
  #---------
  labs(title = "Distribution of Age by Gender", x = "Age", y = "Count",) +
  theme(plot.title = element_text(hjust = 0), 
        axis.title = element_text(size = 10))
```

### 2-C: Age Distribution and Proportion By Decade

```{r}
#| label: Age Distribution and Proportion By Decade
#| fig-width: 14
#| fig-height: 9
#| fig-dpi: 500
#| echo: false
#| message: false
# Barplot of age categories: Colored
####################################

decade1 <- eda_meta_df %>% 
  # Set Up 
  # ------ 
filter (!is.na(age_decade)) %>% 
  # Plotting 
  # -------- 
ggplot(aes(x=age_decade, fill = disease_class)) + 
  geom_bar() + 
  scale_fill_manual(values = progression_colors, 
                    name = "CRC Progression \nClassifiers") + 
  # Labeling 
  #---------
  labs(
    title = "Age Distribution By Decade", 
    x = "Age by Decade", y = "Count")

# Proportion: Age Decade
###########################
decade2 <- eda_meta_df %>%
  # Set Up
  # ------
filter (!is.na(age_decade)) %>%
  # Plotting
  # --------
ggplot(aes(x=age_decade, fill = disease_class)) +
  geom_bar(position="fill") +
  scale_fill_manual(values = progression_colors, 
                    name = "CRC Progression \nClassifiers") +
  # Labeling
  #---------
labs(title = "CRC Progression Proportion by Age Decade",
     x = "Age By Decade", y = "Proportion")

# Patchwork
(decade1 | decade2) + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

```

### 2-D: Diet Bar Plots

```{r}
#| label: Diet counts and Proportion
#| fig-width: 14
#| fig-height: 9
#| fig-dpi: 500
#| echo: false
#| message: false
# Diet ~~~~~~~~~~~~~~~~~~~~~
############################

# Diet Counts
#############
diet1 <- eda_meta_df %>%
  # Set Up
  # ------
  filter(!is.na(diet)) %>%
  # Plotting
  # --------
  ggplot(aes(x = reorder(diet, diet, 
                         FUN = function(x) -length(x)))) + 
  geom_bar(aes(fill = disease_class)) +
  scale_fill_manual(values = progression_colors, 
                    name = "CRC Progression Classifiers") +
  # Labeling
  #---------
  geom_text(stat = "count", aes(label = after_stat(count)), 
            vjust = -0.17, color = "black") +
  ggtitle("Diet of Individuals") +
  labs(title = "Diet of Individuals",
       x = "Diet", 
       y = "Counts")

# Diet Proportion
#################
diet2 <- eda_meta_df %>%
  # Set Up
  # ------
  filter(!is.na(diet)) %>%
  # Plotting
  # --------
  ggplot(aes(x = reorder(diet, diet, 
                         FUN = function(x) -length(x)))) + 
  geom_bar(aes(fill = disease_class), position = "fill") +
  scale_fill_manual(values = progression_colors, 
                    name = "CRC Progression Classifiers") +
  # Labeling
  #---------
  labs(
    title = "Diet of Individuals",
    x = "Diet", 
    y = "Proportion")

# Patchwork
###########
(diet1 | diet2) + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

### 2-E: Country Distribution and Proportions

```{r}
#| label: Countries Associations
#| fig-width: 14
#| fig-height: 6
#| fig-dpi: 500
#| echo: false
#| message: false
# ---

# Countries ~~~~~~~~~~~~~~~~
############################

# Country Count
###############
country1 <- eda_meta_df  %>% 
  # Set Up
  # ------
  filter(!is.na(country)) %>%
  # Plotting
  # --------
  ggplot(aes(x = reorder(country, country, 
                         FUN = function(x) -length(x)))) + 
  geom_bar() +
  # Labeling
  #---------
  geom_text(stat = "count", aes(label = after_stat(count)), 
            vjust = -0.21, size = 3, color = "black") +
  labs(
    title = "Country Sample Count Distribution", 
    x = "Country (NATO country code, trigram)", 
    y = "Count") +
  theme( 
        legend.position = "none", 
        plot.title = element_text(size = 12))

# Country Fill
##############
country2 <- eda_meta_df  %>% 
  # Set Up
  # ------
  filter(!is.na(country)) %>%
  # Plotting
  # --------
  ggplot(aes(x = reorder(country, country, 
                         FUN = function(x) -length(x)), 
             fill=disease_class)) + 
  geom_bar() +
  scale_fill_manual(values = progression_colors, 
                    name = "CRC Progression Classifiers") +
  # Labeling
  #---------
  labs(title = "Country Sample Count Distribution",
       subtitle = "with CRC Progression Classifiers",
       x = "Country (NATO country code, trigram)", 
       y = "Count") +
  theme( 
        legend.position = "none", 
        plot.title = element_text(size = 12))

# Countries Proportions
#######################
country3 <- eda_meta_df %>% 
  # Set Up
  # ------
  filter(!is.na(country)) %>%
  # Plotting
  # --------
  ggplot(aes(x = reorder(country, country, 
                         FUN = function(x) -length(x)), fill = disease_class)) + 
  geom_bar(position = "fill") +
  scale_fill_manual(values = progression_colors, 
                    name = "CRC Progression Classifiers") +
  # Labeling
  #---------
  labs(
    title = "Proportion of CRC Progression Per Country", 
    x = "Country (NATO country code, trigram)", 
    y = "Proportion") +
  theme( 
        plot.title = element_text(size = 12))


# Display
#########
layout3 <- c( # t, l , b = t, r = l
  area(1, 1, 2, 2),
  area(1, 3, 2, 4),
  area(1, 5, 2, 6)
)

#plot(layout3)

geography_caption <- str_wrap(
  "(A) Bar plot of country sample count, sorted in descending order. 
   (B) Stacked bar plot of country sample count, sorted in descending order, filled with CRC progression classifiers.
   (C) Stacked bar plot of the CRC progression classifier proportion for each country in the multi-cohort study.", 
  width = 200
)

country1 + country2 + country3 + 
  plot_layout(design = layout3, guides = "collect", heights = c(1, 0.18)) +
  plot_annotation(
    tag_levels = "A",
    caption = geography_caption
  ) &
  theme(
        plot.title = element_text(hjust = 0, size = 14),
        plot.caption = element_text(hjust = 0, size = 11), 
        legend.key.size = unit(0.5, "cm"),
        legend.title = element_text(size = 10), 
        legend.direction = "vertical", 
        legend.position = "right"
  )
```

### 2-F: Contributors and CRC Progression Classifier Counts

```{r}
#| label: Contributors and CRC Progression Classifier Counts
#| echo: false
#| message: false
# Bar Chart
###########
contributor1 <- eda_meta_df %>%
  # Plotting
  # --------
  ggplot(aes(x = reorder(study_name, study_name, 
                         FUN = function(x) -length(x)),
                         fill = disease_class)) +
  geom_bar() +
  scale_fill_manual(values = progression_colors, 
                    name = "CRC Progression \nClassifiers") +
  theme(axis.text.x = element_text(
    angle = 20, hjust = 1, lineheight = 1.1, size = 5), 
    plot.title = element_text(size = 10), 
    axis.title.x = element_text(size = 10)) +
  # Labeling
  #---------
  xlab("Author Last Name, First Initial, and Year") +
  ylab("Sample Count") +
  ggtitle("Qualifying Study Sample Count \nby CRC Progression Class")

# Bar Chart by Study
####################
contributor2 <- eda_meta_df %>%
  # Plotting
  # --------
  ggplot(aes(x = reorder(study_name, study_name, 
                         FUN = function(x) -length(x)),
                         fill = disease_class)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = progression_colors, 
                    name = "CRC Progression \nClassifiers") +
  theme(axis.text.x = element_text(
    angle = 20, hjust = 1, lineheight = 1.1, size = 5), 
    plot.title = element_text(size = 10), 
    axis.title.x = element_text(size = 10)) +
  # Labeling
  #---------
  xlab("Author Last Name, First Initial, and Year") +
  ylab("Proportion") +
  ggtitle("Qualifying Study Sample Proportions \nby CRC Progression Class")

# Bar Chart of Total CRC Progression Class Proportions
#########################################################
class_count <- eda_meta_df %>%
  # Plotting
  # --------
  ggplot(aes(x=disease_class, stat = "count", fill = disease_class)) +
  geom_bar() +
  scale_fill_manual(values = progression_colors, 
                    name = "CRC Progression \nClassifiers") +
  theme(axis.text.x = element_text(
    angle = 20, hjust = 1, lineheight = 1.1, size = 8), 
    plot.title = element_text(size = 10))+
  # Labeling
  #---------
  labs(y = "Sample Count", x = "CRC Progression Classes") +
  ggtitle("Counts of CRC Progression Classes in Study")

class_count + (contributor1 / contributor2)  + plot_layout(guides = "collect") &
  theme( 
        legend.key.size = unit(0.4, "cm"), 
        legend.title = element_text(size = 12)
  )
```

## 3 - Statistical Analyses

### 3-A: Gender Binomial Test

```{r}
#| label: Gender Binomial Test
#| message: false

sex_counts <- table(eda_meta_df$gender)
binom.test(sex_counts["male"],
           sum(sex_counts),
           p = 0.5) # assumes 50/50 split in population
```

### 3-B: CRC Progression Classifier and Age ANOVA

```{r}
#| label: Age vs CRC Progression ANOVA
#| message: false

age_CRC_ANOVA <- eda_meta_df %>%
  filter(!is.na(age)) 

age_CRC_ANOVA2 <- aov(age ~ disease_class, data = age_CRC_ANOVA)
  
summary(age_CRC_ANOVA2)

TukeyHSD(age_CRC_ANOVA2, conf.level=.95)
```

### 3-C: CRC Progression Classifier and BMI ANOVA

```{r}
#| label: BMI vs CRC Progression ANOVA
#| message: false

bmi_CRC_ANOVA <- eda_meta_df %>%
  filter(!is.na(BMI)) 

bmi_CRC_ANOVA2 <- aov(BMI ~ disease_class, data = bmi_CRC_ANOVA)
  
summary(bmi_CRC_ANOVA2)

TukeyHSD(bmi_CRC_ANOVA2, conf.level=.95)
```

### 3-D: Linear Regression Model by CRC Progression Classifier

```{r}
#| label: LM to match faceted scatterplot
#| message: false

eda_meta_df$age_centered <- scale(
  eda_meta_df$age,
  center = TRUE,
  scale = FALSE
)

eda_meta_df$disease_class <- relevel(
  eda_meta_df$disease_class,
  ref = "HC" # Changed to refer to healthy control rather than "Other"
)

age_bmi_scatterplot_lm <- lm(BMI ~ age_centered * disease_class, 
               data = eda_meta_df)

summary(age_bmi_scatterplot_lm)

```

## 4 - Relevant R Scripts

### 4-A: cMD3 Installation

A script used to install BiocManager and the cMD3 package for the first time.

```{r}
#| label: Installing cMD3
#| eval: false

if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("curatedMetagenomicData")
```

### 4-B: Dry_Runs.R

A script used to determine what information was available on our qualifying studies and print to individual text files. Text files can be found on the GitHub for this project in the described file path in the script.

```{r}
#| label: Dry Runs
#| eval: false

# Run Qualifying Studies List
source("./EDA/eda_meta_df_List.R")

# Studies available
###################
studies <- eda_meta_df %>%
  distinct(study_name) %>%
  pull(study_name)
print(studies)

# Information Available per Study
#################################
for (study in studies) {
  outfile <- paste0("./EDA/Kaitlyn_EDA/Info_Availability/", study, "_info_availability.txt")
  writeLines(curatedMetagenomicData(study, dryrun = TRUE), outfile)
}

# Dry Run
#########
Test <- curatedMetagenomicData("ThomasAM_2018b.+", 
                               dryrun = TRUE) # Doesn't download the data, dry runs it
class(Test)
length(Test)
head(Test)
```

### 4-C: Group EDA Setup R Chunk

The R chunk at the beginning of the file to allow for rendering of all plots and data analysis.

```{r}
#| label: Setup for EDA Document
#| eval: false
source("eda_meta_df_List.R")
suppressPackageStartupMessages({
  library(tidyverse)
  library(knitr)
  library(kableExtra)
  library(patchwork)

  library(vegan)
  library(mia)
  library(scater)
  library(lefser)

  library(grid)
  library(gridExtra)
})

progression_order <- c(
  "HC",
  "PA",
  "PA+",
  "PA-M",
  "CRC",
  "CRC+",
  "CRC-H",
  "CRC-M", 
  "Other"
)

progression_colors <- c(
  "HC" = "#539deb",
  "PA" = "#e4e85b",
  "PA+" = "#e8a25b",
  "CRC" = "#cf1919",
  "CRC+" = "#8f0000",
  "CRC-M" = "#4d0404",
  "CRC-H" = "#520f76",
  "Other" = "#aaa3a3"
)

eda_meta_df <- eda_meta_df %>% 
  mutate(
    disease_class = factor(disease_class, levels = progression_order)
  )
```

## 5 - Relevant Links

### 5-A: Project GitHub

Link to our project GitHub Repository: <https://github.com/chejj/Therayess>

### 5-B: cMD3 Data Set

Link to the cMD3 data set Bioconductor Page: <https://bioconductor.org/packages/3.16/data/experiment/html/curatedMetagenomicData.html>

## 6 - Packages

```{r}
citation("tidyverse")
citation("knitr")
citation("kableExtra")
citation("patchwork")  

citation("vegan")
citation("mia")
citation("scater")
citation("lefser")

citation("grid")
citation("gridExtra")
```
